Hot and fresh: evidence of climate-related suboptimal conditions for seagrass in a large Gulf coast estuary

### 2024-02-08 manu-draft.qmd
reproc all data with sal mid replace for 1975

Linear mixed-effects models for the EPC data showed similar trends in each bay segment for the number of days when temperature was above different thresholds. All of the temperature models for each of the three thresholds (29, 30, 31 $^\circ$C) showed increasing trends for each bay segment, with the largest slope of 1.4 days per year in OTB when temperature was above 29 $^\circ$C (Table S3). The estimated slopes for the number of days when temperature was above 30 $^\circ$C varied from 0.9 days per year in LTB to 1.1 days per year in MTB. Likewise, the mean number of days when temperature was above 30 $^\circ$C at the beginning and end of the period of record varied from 8 (OTB, HB) to 21 (LTB) days in 1975 to 55 (HB) to 63 (LTB) days in 2022 (Table S5, see Table S6 for 1998 to 2022). Trends for the other two temperature thresholds were similar.

@@

An additional caveat of our models was the use of "thresholds" to define potential stressor metrics for temperature and salinity on an annual time scale. Our choice to use 30 $^\circ$C and 25 ppt for temperature and salinity was primarily a statistical consideration given a consistent increase over time in the number of days when these thresholds were crossed. That is, sufficient change and variation in the independent variables for the models of seagrass change were needed to statistically describe potential relationships. These thresholds are likely not ecologically significant since most of the reported values in tropical and sub-tropical environments suggest that the limits of the ecological niche for seagrasses are higher for temperature and lower for salinity [@phillips1960; @mcmillan1967; @zieman1975; @lirman2003]. Because we did not see a dramatic increase in the number of days each year when the thresholds were crossed at more stressful values, conditions in Tampa Bay in recent years are generally within the ecological niche for seagrasses. This does not suggest that these factors are unimportant, both currently and in the future. Extreme temperature or precipitation events acting individually or in combination are likely captured by the trends in stressor metrics using these conservative thresholds, i.e., an increase in a bay segment median number of days also suggests extremes are increasing given the variation around these summary metrics (@fig-mixeff). Further, our models suggested that temperature and salinity are at least associated with seagrass loss and, if so, long-term trends in both are set to amplify their effect.

>>>>

Linear mixed-effects models for the EPC data showed similar trends in each bay segment for the number of days when temperature was above different thresholds. All the temperature models for each of the three thresholds (29, 30, 31 $^\circ$C) showed increasing trends for each bay segment, with the largest slope of 1.4 days per year in OTB when temperature was above 29 $^\circ$C (Table S3). The estimated slopes for the number of days when temperature was above 30 $^\circ$C varied from 0.9 days per year in LTB to 1.1 days per year in MTB. Likewise, the mean number of days when temperature was above 30 $^\circ$C at the beginning and end of the period of record varied from 8 (OTB, HB) to 21 (LTB) days in 1975 to 55 (HB) to 63 (LTB) days in 2022 (Table S5, see Table S6 for 1998 to 2022). Trends for the other two temperature thresholds were similar.

@@

An additional caveat of our models was the use of "thresholds" to define potential stressor metrics for temperature and salinity on an annual time scale. Our choice to use 30 $^\circ$C and 25 ppt for temperature and salinity was primarily a statistical consideration given a consistent increase over time in the number of days when these thresholds were crossed. That is, sufficient change and variation in the independent variables for the models of seagrass change were needed to statistically describe potential relationships. The reported threshold values in tropical and sub-tropical environments suggest that the limits of the ecological niche for seagrasses are higher for temperature and lower for salinity [@phillips1960; @mcmillan1967; @zieman1975; @lirman2003]. Because we did not see a dramatic increase in the number of days each year when the thresholds were crossed at more stressful values, conditions in Tampa Bay in recent years are generally within the ecological niche for seagrasses. This does not suggest that these factors are unimportant, both currently and in the future. Extreme temperature or precipitation events acting individually or in combination are likely captured by the trends in stressor metrics using these lower thresholds, i.e., an increase in a bay segment median number of days also suggests extremes are increasing given the variation around these summary metrics (@fig-mixeff). Our thresholds may also be indicative of the potential for chronic sublethal effects of stress on seagrasses, reducing their resilience to other stressors. Regardless, our models suggested that temperature and salinity are at least associated with seagrass loss and, if so, long-term trends in both are set to amplify their effect.
